Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-03 paper.tex
trigger action
Building models and complexity

% lorem ipsum dummy text
\usepackage{lipsum}

@@

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron P. Ragsdale}
\author[2]{Timothy D. Weaver}
\author[3]{Elizabeth G. Atkinson}
\author[4]{Eileen Hoal}
\author[5]{Marlo M\"{o}ller}
\author[2,6,*]{Brenna M. Henn}
\author[7,**]{Simon Gravel}
\affil[1]{FIXME: UW Madison}
\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA 95616, USA}
\affil[3]{FIXME: DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[4]{FIXME}
\affil[5]{FIXME}
\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA 95616, USA}
\affil[7]{FIXME: McGill}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\maketitle

@@

@@

\section*{Methods}
\cite{Kelleher2016-lw}

>>>>

@@

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron P. Ragsdale}
\author[2]{Timothy D. Weaver}
\author[3]{Elizabeth G. Atkinson}
\author[4]{Eileen Hoal}
\author[5]{Marlo M\"{o}ller}
\author[2,6,*]{Brenna M. Henn}
\author[7,**]{Simon Gravel}
\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
\affil[3]{FIXME: DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[4]{FIXME}
\affil[5]{FIXME}
\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
\affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\maketitle

@@

The intro..

@@

\section*{Methods}
A placeholder citation \citep{Kelleher2016-lw}.

### 2022-01-03 supplement.tex
Building models and complexity

>>>>

With up to six sampled populations in final demographic models that we fit,
there are many parameters to learn. Even in the simplest model involving all
populations (such as the tree-like single-origin model), there are a few dozen
parameters defining split times, migration rates, admixture timings and
proportions, and population sizes and size changes. Thus, parameter space for a
given model topology is large. In addition, the space of possible model
topologies is itself large – as the number of populations increases, the number
of possible topologies also increases, as there are more possibilities for the
order of divergence and admixture events.

In order to narrow the set of possible models to plausible scenarios and to
avoid overfitting, we took an approach that combined the incremental addition
of complexity, starting with fewer populations before combining all
populations, as well as fixing parameters that have been previously estimated
or that fit consistently across all model scenarios. By initially considering
sets of two or three populations, we were able to narrow down the relative
orders of divergences between African and Eurasian populations. Assuming simple
isolation-with-migration models, the Nama appeared to be the earliest diverging
population of those we considered, with West African (Mende) and East African
(Gumuz) populations diverging more recently, followed by the split of the
Eurasian branch from the East African branch.

We performed an initial round of optimization including all six populations
with a family of models including single- and multiple-stem scenarios as
described in the previous section. We identified parameters that reached
consistent values across all models. These included the timings of recent or
non-central divergences and admixture events:
\begin{itemize}
    \item East/West African population split: 60ka
    \item East African/European split: 50 ka
    \item Neanderthal introgression to Eurasian branch: 45 ka
    \item Neanderthal/human split: 550 ka
    \item Eurasian back-to-African migration: 12 ka
\end{itemize}
When testing multiple variations of the more complex models, we kept these
values fixed. This reduced the potential for overfitting the more complex
models, while reducing the computational cost of optimization.

Our models also included recent events to account for known migrations,
admixtures, and growths and declines in effective population sizes. Many of
these parameters were fixed based on previous historical, genetic, or
anthropological research, namely
\begin{itemize}
    \item East African pastoralist to South African admixture: 2 ka \cite{}
    \item European to South African admixture: 10 generations ago \cite{}
    \item Mende population expansion and Gumuz population decline: 5 ka \cite{}
    \item South African population decline following colonial admixture: 9 generations ago \cite{}
\end{itemize}
While the dates of these events were fixed, the sizes and proportions were
allowed to vary in the fits. The total number of parameters that were
ultimately inferred were 18-26 \aprcomment{check!}, depending on the complexity
of the model.
